Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-23 tsibbletalk.Rmd
linking section, still more to do

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to categorical linking where marking one or more observations in one category will broadcast to all other observations in this category. Given time series plots, click any data point on a line, highlighting the whole line as a result. The `as_shared_tsibble()` uses `tsibble`'s `key` variables to achieve these types of linking, and the `spec` argument takes one step further in constructing hybrid linking, for example hierarchical and categorical linking. For example, each series in the `aus_retail` data corresponds to all possible combinations of the `State` and `Industry` variables. They are intrinsically crossed with each other. If one variable is nested within another, this lends itself to a hierarchical structure, like geographical hierarchy. Such collection of inter-related time series are referred to as hierarchical and grouped time series in the literature [@fpp].

To incorporate structured specifications in the `key`, a symbolic formula can be passed to the `spec` argument. Adopting Wilkinson notations for factorial models [@Wilkinson1973], the `spec` follows the `/` and `*` operators tradition to declare nesting and crossing variables respectively. The `spec` for the `aus_retail` data is therefore specified as `State * Industry` or `Industry * State`, which is the default for the presence of multiple `key` variables. If there is a hierarchy in the data, using `/` is required to indicate the parent-child relation, as strictly one direction `parent/child`.

The `tourism_monthly` dataset [@tourism] packaged in \CRANpkg{tsibbletalk}, contains monthly domestic overnight trips across Australia, to give an illustrator of nesting and crossing. The `key` is comprised of three identifying variables: `State`, `Region`, and `Purpose` (of trip), in particular `State` nesting of `Region`, together crossed with `Purpose`. This specification can be translated as follows:

@@

This dataset contains a three-level hierarchy: the root node is implicitly Australia, and geographically disaggregated to states and lower-level tourism regions. A new handy function `plotly_key_tree()` has been implemented to address the need of hierarchical discovery arising from the data. It interprets hierarchies in the shared tsibble's `spec` as a tree view, built with \CRANpkg{plotly}. The following code line produces the linked tree diagram and spans the left panel of Figure \ref{fig:tourism-linking-fig}. The visual of tree hierarchy untangles a group of related series and snapshots the data organisation from a bird's eye view.

>>>>

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to categorical variable linking where selecting one or more observations in one category will broadcast to all other observations in this category. In addition, linking is also by time: within the time series plot, click on any data point, and  whole line will be highlighted in response. 
The `as_shared_tsibble()` uses `tsibble`'s `key` variables to achieve these types of linking.

The approach can also accommodate hierarchical temporal data. When one key variable is nested within another, such as regional areas within a state, this is considered to be a hierarchical structure. These time series are referred to as hierarchical and grouped time series in the literature [@fpp].  The `aus_retail` above is an example. Each series in the  data corresponds to all possible combinations of the `State` and `Industry` variables, which means they are intrinsically crossed with each other.

The `spec` argument in `as_shared_tsibble()`  constructing hybrid linking, that incorporates hierarchical and categorical linking. A symbolic formula can be passed to the `spec` argument, to incorporate structured specifications in the `key`. Adopting @Wilkinson1973's notation for factorial models, the `spec` follows the `/` and `*` operator conventions to declare nesting and crossing variables, respectively. The `spec` for the `aus_retail` data is therefore specified as `State * Industry` or `Industry * State`, which is the default for the presence of multiple `key` variables. If there is a hierarchy in the data, using `/` is required to indicate the parent-child relation, for a strictly one directional `parent/child`.

To illustrate nesting and crossing we use the `tourism_monthly` dataset [@tourism] packaged in \CRANpkg{tsibbletalk}. It contains monthly domestic overnight trips across Australia. The `key` is comprised of three identifying variables: `State`, `Region`, and `Purpose` (of trip), in particular `State` nesting of `Region`, together crossed with `Purpose`. This specification can be translated as follows:

@@

There is a three-level hierarchy: the root node is implicitly Australia, and geographically disaggregated to states and lower-level tourism regions. A new handy function `plotly_key_tree()` has been implemented to help explore the hierarchy. It interprets hierarchies in the shared tsibble's `spec` as a tree view, built with \CRANpkg{plotly}. The following code line produces the linked tree diagram (left panel of Figure \ref{fig:tourism-linking-fig}). The visual for the tree hierarchy detangles a group of related series and provides a bird's eye view of the data organisation.

### 2020-09-23 tsibbletalk.tex
linking section, still more to do

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to
categorical linking where marking one or more observations in one
category will broadcast to all other observations in this category.
Given time series plots, click any data point on a line, highlighting
the whole line as a result. The \texttt{as\_shared\_tsibble()} uses
\texttt{tsibble}'s \texttt{key} variables to achieve these types of
linking, and the \texttt{spec} argument takes one step further in
constructing hybrid linking, for example hierarchical and categorical
linking. For example, each series in the \texttt{aus\_retail} data
corresponds to all possible combinations of the \texttt{State} and
\texttt{Industry} variables. They are intrinsically crossed with each
other. If one variable is nested within another, this lends itself to a
hierarchical structure, like geographical hierarchy. Such collection of
inter-related time series are referred to as hierarchical and grouped
time series in the literature \citep{fpp}.

To incorporate structured specifications in the \texttt{key}, a symbolic
formula can be passed to the \texttt{spec} argument. Adopting Wilkinson
notations for factorial models \citep{Wilkinson1973}, the \texttt{spec}
follows the \texttt{/} and \texttt{*} operators tradition to declare
nesting and crossing variables respectively. The \texttt{spec} for the
\texttt{aus\_retail} data is therefore specified as
\texttt{State\ *\ Industry} or \texttt{Industry\ *\ State}, which is the
default for the presence of multiple \texttt{key} variables. If there is
a hierarchy in the data, using \texttt{/} is required to indicate the
parent-child relation, as strictly one direction \texttt{parent/child}.

The \texttt{tourism\_monthly} dataset \citep{tourism} packaged in
\CRANpkg{tsibbletalk}, contains monthly domestic overnight trips across
Australia, to give an illustrator of nesting and crossing. The
\texttt{key} is comprised of three identifying variables:
\texttt{State}, \texttt{Region}, and \texttt{Purpose} (of trip), in
particular \texttt{State} nesting of \texttt{Region}, together crossed
with \texttt{Purpose}. This specification can be translated as follows:

@@

This dataset contains a three-level hierarchy: the root node is
implicitly Australia, and geographically disaggregated to states and
lower-level tourism regions. A new handy function
\texttt{plotly\_key\_tree()} has been implemented to address the need of
hierarchical discovery arising from the data. It interprets hierarchies
in the shared tsibble's \texttt{spec} as a tree view, built with
\CRANpkg{plotly}. The following code line produces the linked tree
diagram and spans the left panel of Figure
\ref{fig:tourism-linking-fig}. The visual of tree hierarchy untangles a
group of related series and snapshots the data organisation from a
bird's eye view.

>>>>

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to
categorical variable linking where selecting one or more observations in
one category will broadcast to all other observations in this category.
In addition, linking is also by time: within the time series plot, click
on any data point, and whole line will be highlighted in response. The
\texttt{as\_shared\_tsibble()} uses \texttt{tsibble}'s \texttt{key}
variables to achieve these types of linking.

The approach can also accommodate hierarchical temporal data. When one
key variable is nested within another, such as regional areas within a
state, this is considered to be a hierarchical structure. These time
series are referred to as hierarchical and grouped time series in the
literature \citep{fpp}. The \texttt{aus\_retail} above is an example.
Each series in the data corresponds to all possible combinations of the
\texttt{State} and \texttt{Industry} variables, which means they are
intrinsically crossed with each other.

The \texttt{spec} argument in \texttt{as\_shared\_tsibble()}
constructing hybrid linking, that incorporates hierarchical and
categorical linking. A symbolic formula can be passed to the
\texttt{spec} argument, to incorporate structured specifications in the
\texttt{key}. Adopting \citet{Wilkinson1973}'s notation for factorial
models, the \texttt{spec} follows the \texttt{/} and \texttt{*} operator
conventions to declare nesting and crossing variables, respectively. The
\texttt{spec} for the \texttt{aus\_retail} data is therefore specified
as \texttt{State\ *\ Industry} or \texttt{Industry\ *\ State}, which is
the default for the presence of multiple \texttt{key} variables. If
there is a hierarchy in the data, using \texttt{/} is required to
indicate the parent-child relation, for a strictly one directional
\texttt{parent/child}.

To illustrate nesting and crossing we use the \texttt{tourism\_monthly}
dataset \citep{tourism} packaged in \CRANpkg{tsibbletalk}. It contains
monthly domestic overnight trips across Australia. The \texttt{key} is
comprised of three identifying variables: \texttt{State},
\texttt{Region}, and \texttt{Purpose} (of trip), in particular
\texttt{State} nesting of \texttt{Region}, together crossed with
\texttt{Purpose}. This specification can be translated as follows:

@@

There is a three-level hierarchy: the root node is implicitly Australia,
and geographically disaggregated to states and lower-level tourism
regions. A new handy function \texttt{plotly\_key\_tree()} has been
implemented to help explore the hierarchy. It interprets hierarchies in
the shared tsibble's \texttt{spec} as a tree view, built with
\CRANpkg{plotly}. The following code line produces the linked tree
diagram (left panel of Figure \ref{fig:tourism-linking-fig}). The visual
for the tree hierarchy detangles a group of related series and provides
a bird's eye view of the data organisation.
